---
title: "Loops and Functions Examples"
output: html_notebook
---

```{r}
library(tidyverse)
```


Basic for loop:
```{r}
for (letter in c("a", "b", "c")){
  print(letter)
}
```

For loop with sequence of numbers:
```{r}
for (i in 5:8) {
  print(i)
}
```

A basic use of map:
```{r}
map(c(-1, 12, 0), abs)
```

Specifying the output type:
```{r}
map_dbl(c(-1, 12, 0), abs)
```

Mapping with a dataframe:
```{r}
penguins <- read_csv("../data/penguins.csv")
```

```{r}
map_int(penguins, n_distinct)
```

With an anonymous function:
```{r}
map_int(penguins, \(x) sum(is.na(x)))
```

Or using a tilde:
```{r}
map_int(penguins, ~sum(is.na(.)))
```

A basic function:
```{r}
square <- function(x){
  x^2
}
```

```{r}
square(10)
```

A more complicated function:
```{r}
collatz <- function(n){
  if (n %% 2 == 0){
    n / 2
  }
  else {
    3*n + 1
  }
}
```

```{r}
collatz(10)
```

```{r}
collatz(13)
```

Syntax for a for loop and the equivalent of an f-string.
```{r}
starting_value = 13
num_iter = 0

n <- starting_value
while (n != 1){
  n <- collatz(n)
  num_iter = num_iter + 1
}

print(glue::glue("{starting_value} required {num_iter} iterations to reach 1."))
```

Functions with tidyverse verbs. The incorrect way:
```{r}
find_mean <- function(summary_variable){
  penguins |> 
    summarize(mean(summary_variable, na.rm = TRUE))
}
```

```{r}
find_mean(bill_length_mm)
```

The correct way - embracing the variable:
```{r}
find_mean <- function(summary_variable){
  penguins |> 
    summarize(mean({{ summary_variable }}, na.rm = TRUE))
}
```

```{r}
find_mean(bill_length_mm)
```

Or, if we wanted to make a nicer column name:
```{r}
find_mean <- function(summary_variable){

  penguins |> 
    summarize("mean_{{ summary_variable }}" := mean({{ summary_variable }}, na.rm = TRUE))
}
```

```{r}
find_mean(bill_length_mm)
```

