---
title: "Working with Geospatial Data in R"
output: html_notebook
---

```{r}
library(sf)
library(tidyverse)
```

```{r}
zipcodes <- st_read('../data/zipcodes.geojson')
```

We can create plots out of our geospatial data using ggplot. 
```{r}
zipcodes |> 
  ggplot() +
  geom_sf()
```

We can adjust the color using the fill aesthethic.
```{r}
zipcodes |> 
  ggplot() +
  geom_sf(aes(fill = poname))
```

```{r}
bus_stops <- read_csv("../data/WeGo_Transit_Bus_Stops_20231025.csv")
```

Let's extract the latitude and longitude using the str_extract function from stringr.
```{r}
bus_stops <- bus_stops |> 
  mutate(latitude = as.numeric(str_extract(`Mapped Location`, "\\((.+),", group = 1)),
            longitude = as.numeric(str_extract(`Mapped Location`, ",(.+)\\)", group = 1))
            )
```

```{r}
zipcodes |> 
  ggplot() +
  geom_sf() +
  geom_point(data = bus_stops |> drop_na(latitude),
             aes(x = longitude, y = latitude),
             size = 0.1)
```

If we want to be able to perform geospatial operations on our bus stops, we need to convert it into a geospatial object. We can do this using the st_as_sf function.
```{r}
bus_stops_geo <- st_as_sf(bus_stops |> drop_na(latitude),
         coords = c('longitude', 'latitude'),
         crs = st_crs(zipcodes)
         )
```

```{r}
zipcodes |> 
  ggplot() +
  geom_sf() +
  geom_sf(data = bus_stops_geo, size = 0.1)
```

If we want to perform spatial joins, the sf library has an st_join function. We need to specify how to join (contains, intersects, etc.). Here, we'll use containment.

```{r}
bus_zips <- st_join(bus_stops_geo, zipcodes, join = st_within)
```

Notice that bus_zips retains the geometry of bus_stops, as this was the first dataframe we provided.
Now, we can, for example, count the number of stops per zipcode. Note that when performing calculations on an sf object, it is usually quicker if you drop the geometry first.
```{r}
bus_zips |> 
  st_drop_geometry() |> 
  group_by(zipcode) |> 
  count(name = "num_stops") |> 
  arrange(desc(num_stops))
```

Now, we can plot the bus stops contained in a given zipcode.
```{r}
zip = 37207

zipcodes |> 
  filter(zipcode == zip) |> 
  ggplot() +
  geom_sf() +
  geom_sf(data = bus_zips |> filter(zipcode == zip),
          aes(color = `Route Name`))
```

Finally, if we want to make an interactive map, we can use the leaflet library (https://rstudio.github.io/leaflet/).
```{r}
library(leaflet)
```

```{r}
leaflet(data = bus_stops |> 
          drop_na(latitude) |> 
          filter(`Route Name` == "WEST END - WHITE BRIDGE")
        ) |>  
  addTiles() |> 
  addMarkers(~longitude, 
             ~latitude, 
             popup = ~as.character(`Stop Name`), 
             label = ~as.character(`Stop Name`)
             )
```

You can also do marker clusters.
```{r}
leaflet(bus_stops |> drop_na(longitude)) %>% 
  addTiles() %>% 
  addMarkers(
    ~longitude,
    ~latitude,
    clusterOptions = markerClusterOptions()
  )
```

