---
title: "Principal Components Analysis (PCA)"
output:
  html_document:
    df_print: paged
---

In this notebook, we will explore principal components analysis using the iris dataset. This dataset consists of measurements of 4 components of 3 different species of iris and is included in base R.
```{r}
head(iris)
```

Let's create some plots to get an understanding of our dataset. We can start by picking two variables and creating a scatterplot. We'll color by species to see how well we can distinguish the species on the basis of these measurements.

```{r}
library(tidyverse)
```

```{r}
iris %>% 
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color = Species)) + geom_point()
```

Let's try another pair of variables:
```{r}
iris %>% 
  ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species)) + geom_point()
```

And another:
```{r}
iris %>% 
  ggplot(aes(x = Sepal.Length, y = Petal.Length, color = Species)) + geom_point()
```

By changing the pair of variables we chose, we get different views of the dataset. But did we choose the "best" pair of variables to understand our dataset. What if we had a few hundred variables? It would be time-consuming to try and sort through all possible pairs. Luckily, there is a tool that can be used to find the "best" 2D scatterplot for our data. The catch is that our axes may not consist of our original variables, but instead linear combinations of the original variables.

We will be using the prcomp function from the stats library (which is automatically loaded). This function computes the principal components of our dataset. 

When doing PCA, you almost always want your variables to be on "equal footing". That is, you don't the variable with the highest variance to dominate the others. 

When using the prcomp function, we can specify that we would like to center and scale our data.

```{r}
iris.pr <- prcomp(iris %>% select(-Species), center = TRUE, scale = TRUE)
summary(iris.pr)
```

The summary shows that the first pricipal component captures 72.96% of the variance in the dataset, and that the first two components combined capture 95.81%. This means that if we project our datset onto the plane generated by these two components, we will get a pretty good picture of the distribution of our datapoints (only missing out on about 4% of the variance).

Let's see what this projection looks like and how well we can distinguish the species.
```{r}
iris_projection <- iris.pr$x %>% 
  as_tibble() %>% 
  cbind(iris$Species) %>% 
  rename(Species = `iris$Species`)
iris_projection %>% 
  ggplot(aes(x = PC1, y = PC2, color = Species)) + geom_point()
```

To understand how much each original variable contributes to the principal components, we can inspect the rotation matrix. Important note: these contributions are with respect to the centered and scaled values of the original variables!
```{r}
iris.pr$rotation
```
This says to find the projection onto the first prinicipal component, the formula is 
$$0.5210659*(\text{Scaled Sepal.Length}) - 0.2693474*(\text{Scaled Sepal.Width}) + 0.5804131*(\text{Scaled Petal.Length}) + 0.5648565*(\text{Scaled Petal.Width})$$

Lets confirm this by multiplying (matrix multiplication, which is done using %*%) the rotation matrix by the first row.
```{r}
iris_scaled = scale(iris[,1:4])
iris_scaled[1,] %*% iris.pr$rotation
```

Compare to the projection values.
```{r}
iris_projection[1,]
```

What we can see is that the first principal component is determined by an approximately equal mix of sepal length, petal length, and petal width.
```{r}
tibble(coefficient = iris.pr$rotation[,1], variable = names(iris.pr$rotation[,1])) %>% 
  ggplot(aes(x = variable, y = coefficient)) + geom_col() + coord_flip()
```

Whereas, the second pricipal component is mostly determined by the sepal width.
```{r}
tibble(coefficient = iris.pr$rotation[,2], variable = names(iris.pr$rotation[,1])) %>% 
  ggplot(aes(x = variable, y = coefficient)) + geom_col() + coord_flip()
```
