---
title: "Principal Components Analysis (PCA)"
output:
  html_document:
    df_print: paged
---

In this notebook, we will explore principal components analysis using the iris dataset. This dataset consists of measurements of 4 components of 3 different species of iris and is included in base R.
```{r}
head(iris)
```

Let's create some plots to get an understanding of our dataset. We can start by picking two variables and creating a scatterplot. We'll color by species to see how well we can distinguish the species on the basis of these measurements.

```{r}
library(tidyverse)
```

Let's first look at the Petal.Length and Petal.Width variables.
```{r}
iris %>% 
  ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
  geom_point()
```
It appears that these variables are strongly correlated. Let's check it. 
```{r}
iris %>% 
  select(Petal.Length, Petal.Width) %>% 
  cor()
```

In a way, Petal.Length and Petal.Width are redundant. Since they are so correlated, we don't really need both in order to understand the dataset.

One option is to discard one of them. However, in doing so, we will lose out on some information since they are not completely correlated, but only strongly so. If we were to discard the Petal.Width variable, it would be like we were projecting onto a line (determined by Petal.Length).
```{r}
iris %>% 
  ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
  geom_point() +
  geom_hline(yintercept = mean(iris$Petal.Width)) +
  geom_segment(aes(x = Petal.Length, y = Petal.Width, xend = Petal.Length, yend = mean(Petal.Width)))
```

```{r}
iris %>% 
  ggplot(aes(color = Species)) + 
  geom_hline(yintercept = mean(iris$Petal.Width)) +
  geom_point(aes(x = Petal.Length, y = mean(Petal.Width)))
```

There are a couple of potential problems with this.

First, the "reconstruction" distance is quite large. Points with large or small values for Petal.Length end up quite far from where they started.
We'll define the error as the distance from the original point to its projection.
```{r}
iris %>% 
  summarize(avg_error = mean(Petal.Width^2))
```


Second, points that started far from each other end up quite close to each other. That is, we are losing some of the information (variance) in the original dataset.

Let's compare the variance in Petal.Length to the total variance in the dataset.
```{r}
100 * var(iris$Petal.Length) / (var(iris$Petal.Length) + var(iris$Petal.Width))
```


Can we do better? An alternative is to combine them in some way some way into a new feature. One option we can try is to mix equal parts of the two variables. This is essentially the same as projecting (the centered points) onto the line $$y = x$$.

```{r}
angle = 45

slope = sin(angle * pi / 180) / cos(angle * pi / 180)

iris_projection <- iris %>% 
    as_tibble() %>% 
    mutate(projection = (Petal.Length - mean(iris$Petal.Length)) * cos(angle * pi / 180)  + (Petal.Width - mean(iris$Petal.Width)) * sin(angle * pi / 180),
           proj_x = mean(iris$Petal.Length) + projection * cos(angle * pi / 180), proj_y = mean(iris$Petal.Width) + projection * sin(angle*pi / 180))

iris_projection %>% 
    ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
    geom_point() +
    geom_abline(slope = slope, intercept = mean(iris$Petal.Width) - slope*mean(iris$Petal.Length)) +
    geom_segment(aes(x = Petal.Length, y = Petal.Width, xend = proj_x, yend = proj_y)) +
    coord_fixed(ratio = 1)
```
Let's compare the reconstruction error to what we had before:
```{r}
iris_projection %>% 
  summarize(avg_error = mean(sqrt((Petal.Length - proj_x)^2 + (Petal.Width - proj_y)^2)))
```

Now, let's see how much of the original variance we captured.
```{r}
iris_projection %>% 
    ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
    geom_abline(slope = slope, intercept = mean(iris$Petal.Width) - slope*mean(iris$Petal.Length)) +
    geom_point(aes(x = mean(iris$Petal.Length) + projection * cos(angle * pi / 180), y = mean(iris$Petal.Width) + projection * sin(angle*pi / 180))) +
    coord_fixed(ratio = 1)
```

```{r}
100 * var(iris_projection$projection) / (var(iris$Petal.Length) + var(iris$Petal.Width))
```
Can we do better? Try adjusting the angle value above to try and find a better solution.



To find the "best" such line, we can make use of the prcomp function.
```{r}
iris_pca <- iris %>% 
  select(Petal.Length, Petal.Width) %>% 
  prcomp()
```


We can see the vector determining this line using the rotation attribute. The vector that we care about is given by PC1.
```{r}
iris_pca$rotation
```

Using a bit of trigonometry, we can find the corresponding angle:
```{r}
angle = 180 / pi * atan(iris_pca$rotation['Petal.Width', 'PC1'] / iris_pca$rotation['Petal.Length', 'PC1'])

angle
```

```{r}
angle = angle

slope = sin(angle * pi / 180) / cos(angle * pi / 180)

iris_projection <- iris %>% 
    as_tibble() %>% 
    mutate(projection = (Petal.Length - mean(iris$Petal.Length)) * cos(angle * pi / 180)  + (Petal.Width - mean(iris$Petal.Width)) * sin(angle * pi / 180),
           proj_x = mean(iris$Petal.Length) + projection * cos(angle * pi / 180), proj_y = mean(iris$Petal.Width) + projection * sin(angle*pi / 180))

iris_projection %>% 
    ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
    geom_point() +
    geom_abline(slope = slope, intercept = mean(iris$Petal.Width) - slope*mean(iris$Petal.Length)) +
    geom_segment(aes(x = Petal.Length, y = Petal.Width, xend = proj_x, yend = proj_y)) +
    coord_fixed(ratio = 1)
```
Let's see what the reconstruction error is:
```{r}
iris_projection %>% 
  summarize(avg_error = mean(sqrt((Petal.Length - proj_x)^2 + (Petal.Width - proj_y)^2)))
```

And the amount of variance captured:
```{r}
100 * var(iris_projection$projection) / (var(iris$Petal.Length) + var(iris$Petal.Width))
```

The interesting thing is that this angle optimizes both quantities!


```{r}
iris_projection %>% 
  ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
  geom_point() +
  geom_segment(aes(x = mean(iris$Petal.Length), y = mean(iris$Petal.Width), xend = mean(iris$Petal.Length) + iris_pca$rotation['Petal.Length', 'PC1'], yend = mean(iris$Petal.Width) + iris_pca$rotation['Petal.Width', 'PC1']), color = 'black', arrow = arrow(ends = "last", length = unit(0.15, "inches"), type = "closed")) +
  coord_fixed(ratio = 1)
```


Our new variables are constructed are given by 
$$ 0.92177 \cdot (Petal.Length - mean(Petal.Length)) + 0.3877188 \cdot (Petal.Width - mean(Petal.Width))$$

To get the angle for this projection, we can do some trigonometry:
```{r}
first_component <- iris_pca$rotation[,'PC1']

angle = atan(first_component['Petal.Width'] / first_component['Petal.Length']) * 180 / pi

angle
```

And we can visualize the projection:
```{r}
slope = sin(angle * pi / 180) / cos(angle * pi / 180)

iris_projection <- iris %>% 
    as_tibble() %>% 
    mutate(projection = (Petal.Length - mean(iris$Petal.Length)) * cos(angle * pi / 180)  + (Petal.Width - mean(iris$Petal.Width)) * sin(angle * pi / 180))

iris_projection %>% 
    ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
    geom_point() +
    geom_abline(slope = slope, intercept = mean(iris$Petal.Width) - slope*mean(iris$Petal.Length)) +
    geom_segment(aes(x = Petal.Length, y = Petal.Width, xend = mean(iris$Petal.Length) + projection * cos(angle * pi / 180), yend = mean(iris$Petal.Width) + projection * sin(angle*pi / 180))) +
    geom_segment(aes(x = mean(iris$Petal.Length), y = mean(iris$Petal.Width), xend = mean(iris$Petal.Length) + iris_pca$rotation['Petal.Length', 'PC1'], yend = mean(iris$Petal.Width) + iris_pca$rotation['Petal.Width', 'PC1']), color = 'black', arrow = arrow(ends = "last", length = unit(0.15, "inches"), type = "closed")) +
    coord_fixed(ratio = 1)
```

The new variable represents the distance of the projection along this line.
```{r}
iris_projection %>% 
    ggplot(aes(x = Petal.Length, y = Petal.Width)) + 
    geom_abline(slope = slope, intercept = mean(iris$Petal.Width) - slope*mean(iris$Petal.Length)) +
    geom_point(aes(x = mean(iris$Petal.Length) + projection * cos(angle * pi / 180), y = mean(iris$Petal.Width) + projection * sin(angle*pi / 180), color = Species)) +
    coord_fixed(ratio = 1)
```

The projection values are contained in the x attribute. Here, we are interested in the PC1 column.
```{r}
iris_pca$x
```

Now, let's look at the summary:
```{r}
summary(iris_pca)
```
The PCA summary shows how much of the original variance in the dataset is captured by our projection. This shows that the PC1 (the projection that we used) captures 99.02% of the variability. Here, variability means the sum of the covariances of the original dataset, which we can see from the covar function.

```{r}
covar <- iris %>% 
  select(Petal.Width, Petal.Length) %>% 
  cov()

covar
```
The total variance is the sum of the diagonal entries.
```{r}
covar[1,1] + covar[2,2]
```

We can access the projection via the x attribute. If we calculate the variance in the projection along PC1 with the original variance, we get that 99.02% figure.
```{r}
var(iris_pca$x[,'PC1']) / (covar[1,1] + covar[2,2])
```

What about PC2? First, let's plot it on our scatterplot.
```{r}
iris_projection %>% 
  ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
  geom_point() +
  geom_segment(aes(x = mean(iris$Petal.Length), y = mean(iris$Petal.Width), xend = mean(iris$Petal.Length) + iris_pca$rotation['Petal.Length', 'PC1'], yend = mean(iris$Petal.Width) + iris_pca$rotation['Petal.Width', 'PC1']), color = 'black', arrow = arrow(ends = "last", length = unit(0.15, "inches"), type = "closed")) +
  geom_segment(aes(x = mean(iris$Petal.Length), y = mean(iris$Petal.Width), xend = mean(iris$Petal.Length) + iris_pca$rotation['Petal.Length', 'PC2'], yend = mean(iris$Petal.Width) + iris_pca$rotation['Petal.Width', 'PC2']), color = 'black', arrow = arrow(ends = "last", length = unit(0.15, "inches"), type = "closed")) +
  coord_fixed(ratio = 1)
```
It turns out that this vector is perpendicular to PC1. Projecting along this vector will account for the remaining variance in our dataset.

We can use PC1 and PC2 as a new basis for the coordinate system.

```{r}
iris_pca$x %>% 
  as_tibble() %>% 
  add_column(Species = iris$Species) %>% 
  ggplot(aes(x = PC1, y = PC2, color = Species)) + 
  geom_point()
```

One nice property about PCA is that our new variables are uncorrelated:
```{r}
iris_pca$x %>% 
  cov()
```


## Full Dataset

Now, let's bring in the other two variables. In the full dataset, we have 4 quantitative variables. We have already seen how it is possible to condense our two variables into just a single variable. What if we want to 

And another:
```{r}
iris %>% 
  ggplot(aes(x = Sepal.Length, y = Petal.Length, color = Species)) + geom_point()
```

By changing the pair of variables we chose, we get different views of the dataset. But did we choose the "best" pair of variables to understand our dataset. What if we had a few hundred variables? It would be time-consuming to try and sort through all possible pairs. Luckily, there is a tool that can be used to find the "best" 2D scatterplot for our data. The catch is that our axes may not consist of our original variables, but instead linear combinations of the original variables.

We will be using the prcomp function from the stats library (which is automatically loaded). This function computes the principal components of our dataset. 

When doing PCA, you almost always want your variables to be on "equal footing". That is, you don't the variable with the highest variance to dominate the others. 

When using the prcomp function, we can specify that we would like to center and scale our data.

```{r}
iris.pr <- prcomp(iris %>% select(-Species), center = TRUE, scale = TRUE)
summary(iris.pr)
```

The summary shows that the first pricipal component captures 72.96% of the variance in the dataset, and that the first two components combined capture 95.81%. This means that if we project our datset onto the plane generated by these two components, we will get a pretty good picture of the distribution of our datapoints (only missing out on about 4% of the variance).

Let's see what this projection looks like and how well we can distinguish the species.
```{r}
iris_projection <- iris.pr$x %>% 
  as_tibble() %>% 
  cbind(iris$Species) %>% 
  rename(Species = `iris$Species`)
iris_projection %>% 
  ggplot(aes(x = PC1, y = PC2, color = Species)) + geom_point()
```

To understand how much each original variable contributes to the principal components, we can inspect the rotation matrix. Important note: these contributions are with respect to the centered and scaled values of the original variables!
```{r}
iris.pr$rotation
```
This says to find the projection onto the first prinicipal component, the formula is 
$$0.5210659*(\text{Scaled Sepal.Length}) - 0.2693474*(\text{Scaled Sepal.Width}) + 0.5804131*(\text{Scaled Petal.Length}) + 0.5648565*(\text{Scaled Petal.Width})$$

Lets confirm this by multiplying (matrix multiplication, which is done using %*%) the rotation matrix by the first row.
```{r}
iris_scaled = scale(iris[,1:4])
iris_scaled[1,] %*% iris.pr$rotation
```

Compare to the projection values.
```{r}
iris_projection[1,]
```

What we can see is that the first principal component is determined by an approximately equal mix of sepal length, petal length, and petal width.
```{r}
tibble(coefficient = iris.pr$rotation[,1], variable = names(iris.pr$rotation[,1])) %>% 
  ggplot(aes(x = variable, y = coefficient)) + geom_col() + coord_flip()
```

Whereas, the second pricipal component is mostly determined by the sepal width.
```{r}
tibble(coefficient = iris.pr$rotation[,2], variable = names(iris.pr$rotation[,1])) %>% 
  ggplot(aes(x = variable, y = coefficient)) + geom_col() + coord_flip()
```
