---
title: "Principal Components Analysis (PCA), Part 1"
output:
  html_document:
    df_print: paged
---

In this notebook, we will explore principal components analysis using the iris dataset. This dataset consists of measurements of 4 components of 3 different species of iris and is included in base R. It is a popular dataset for machine learning examples.
```{r}
head(iris)
```

In this notebook, we'll focus on just two of the variables, the petal length and petal width.

```{r}
library(tidyverse)
```

Let's start by creating a scatterplot of the two variables:
```{r}
iris %>% 
  ggplot(aes(x = Petal.Length, y = Petal.Width, fill = Species)) + 
  geom_point(shape = 21, color = 'black', size = 2) +
  coord_fixed(ratio = 1)
```
It appears that these variables are strongly correlated. Let's see how strong this correlation is.
```{r}
iris %>% 
  select(Petal.Length, Petal.Width) %>% 
  cor()
```

Since they are so strongly correlated, Petal.Length and Petal.Width are almost redundant. If we know, for example, the petal length, then we can make a pretty good guess about the petal width. Can we get away with a simpler representation of our dataset so that we don't need to store both of these variables?

One option is to completely discard one of them. However, in doing so, we will lose out on some information since they are not perfectly correlated, but only strongly so. If we were to discard the Petal.Width variable, it would be like we were projecting onto a horizontal line.
```{r}
iris %>% 
  ggplot(aes(x = Petal.Length, y = Petal.Width, fill = Species)) + 
  geom_hline(yintercept = mean(iris$Petal.Width)) +
  geom_segment(aes(x = Petal.Length, y = Petal.Width, xend = Petal.Length, yend = mean(Petal.Width))) +
  geom_point(shape = 21, color = 'black', size = 2) +
  coord_fixed(ratio = 1)
```

```{r}
iris %>% 
  ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
  geom_hline(yintercept = mean(iris$Petal.Width)) +
  geom_point(aes(x = Petal.Length, y = mean(Petal.Width), fill = Species), shape = 21, color = 'black', size = 2)
```

There are a couple of potential problems with this.

First, the "reconstruction" distance is quite large. Points with large or small values for Petal.Length end up quite far from where they started.
We'll define the reconstruction error of this projection as the distance from the original point to its projection. Note that we are projecting onto the line that goes through the mean of each variable.
```{r}
iris %>% 
  summarize(reconstruction_error = mean((Petal.Width - mean(iris$Petal.Width))^2))
```

A second potential problem is that points that started far from each other end up close to each other. That is, we are losing some of the information (variance) in the original dataset.

Let's compare the variance in Petal.Length (our one variable that remains) to the total variance in the dataset.
```{r}
100 * var(iris$Petal.Length) / (var(iris$Petal.Length) + var(iris$Petal.Width))
```

Can we do better? Maybe we can find a better line to project onto. The data shows an upward trend, so we can try projecting onto the line y = x.

```{r}
# The angle of the line in degrees
angle = 45

# For graphing purposes, we need to know the slope of this line
slope = sin(angle * pi / 180) / cos(angle * pi / 180)

# A bit of linear algebra to find the projections
iris_projection <- iris %>% 
    as_tibble() %>% 
    mutate(projection = (Petal.Length - mean(iris$Petal.Length)) * cos(angle * pi / 180)  + (Petal.Width - mean(iris$Petal.Width)) * sin(angle * pi / 180),
           proj_x = mean(iris$Petal.Length) + projection * cos(angle * pi / 180), proj_y = mean(iris$Petal.Width) + projection * sin(angle*pi / 180))

iris_projection %>% 
    ggplot(aes(x = Petal.Length, y = Petal.Width, fill = Species)) + 
    geom_abline(slope = slope, intercept = mean(iris$Petal.Width) - slope*mean(iris$Petal.Length)) +
    geom_segment(aes(x = Petal.Length, y = Petal.Width, xend = proj_x, yend = proj_y)) +
    geom_point(shape = 21, color = 'black', size = 2) +
    coord_fixed(ratio = 1)
```

```{r}
iris_projection %>% 
    ggplot(aes(x = Petal.Length, y = Petal.Width, color = Species)) + 
    geom_abline(slope = slope, intercept = mean(iris$Petal.Width) - slope*mean(iris$Petal.Length)) +
    geom_point(aes(x = proj_x, y = proj_y, fill = Species), shape = 21, color = 'black', size = 2) +
    coord_fixed(ratio = 1, xlim = c(0, 7))
```

First, let's see how much of the original variance we captured.
```{r}
100 * var(iris_projection$projection) / (var(iris$Petal.Length) + var(iris$Petal.Width))
```
We do a bit better than we did before.


Now, let's compare the reconstruction error to what we had before:
```{r}
iris_projection %>% 
  summarize(reconstruction_error = mean(sqrt((Petal.Length - proj_x)^2 + (Petal.Width - proj_y)^2)))
```
We actually did a bit worse on this metric.

Can we do better? Try adjusting the angle value above to try and find a better solution.



To find the "best" such line, we can make use of the prcomp function.
```{r}
iris_pca <- iris %>% 
  select(Petal.Length, Petal.Width) %>% 
  prcomp()
```

We can see the vector determining this line using the rotation attribute. The vector that we care about is given by PC1.
```{r}
iris_pca$rotation
```

```{r}
iris %>% 
  ggplot(aes(x = Petal.Length, y = Petal.Width, fill = Species)) + 
  geom_point(shape = 21, color = 'black', size = 2) +
  geom_segment(aes(x = mean(Petal.Length), y = mean(Petal.Width), xend = mean(Petal.Length) + iris_pca$rotation['Petal.Length', 'PC1'], yend = mean(Petal.Width) + iris_pca$rotation['Petal.Width', 'PC1']), color = 'black', arrow = arrow(ends = "last", length = unit(0.15, "inches"), type = "closed"), size = 2) +
  coord_fixed(ratio = 1)
```

Using a bit of trigonometry, we can find the corresponding angle:
```{r}
optimal_angle = 180 / pi * atan(iris_pca$rotation['Petal.Width', 'PC1'] / iris_pca$rotation['Petal.Length', 'PC1'])

optimal_angle
```

```{r}
angle = optimal_angle

# For graphing purposes, we need to know the slope of this line
slope = sin(angle * pi / 180) / cos(angle * pi / 180)

# A bit of linear algebra to find the projections
iris_projection <- iris %>% 
    as_tibble() %>% 
    mutate(projection = (Petal.Length - mean(iris$Petal.Length)) * cos(angle * pi / 180)  + (Petal.Width - mean(iris$Petal.Width)) * sin(angle * pi / 180),
           proj_x = mean(iris$Petal.Length) + projection * cos(angle * pi / 180), proj_y = mean(iris$Petal.Width) + projection * sin(angle*pi / 180))

iris_projection %>% 
    ggplot(aes(x = Petal.Length, y = Petal.Width, fill = Species)) + 
    geom_abline(slope = slope, intercept = mean(iris$Petal.Width) - slope*mean(iris$Petal.Length)) +
    geom_segment(aes(x = Petal.Length, y = Petal.Width, xend = proj_x, yend = proj_y)) +
    geom_point(shape = 21, color = 'black', size = 2) +
    coord_fixed(ratio = 1)
```
Let's see what the reconstruction error is:
```{r}
iris_projection %>% 
  summarize(avg_error = mean(sqrt((Petal.Length - proj_x)^2 + (Petal.Width - proj_y)^2)))
```

And the amount of variance captured:
```{r}
100 * var(iris_projection$projection) / (var(iris$Petal.Length) + var(iris$Petal.Width))
```

The interesting thing is that this angle optimizes both quantities!

```{r}
find_error <- function(angle){
  iris_projection <- iris %>%
    mutate(projection = (Petal.Length - mean(iris$Petal.Length)) * cos(angle * pi / 180)  + (Petal.Width - mean(iris$Petal.Width)) * sin(angle * pi / 180),
           proj_x = mean(iris$Petal.Length) + projection * cos(angle * pi / 180), proj_y = mean(iris$Petal.Width) + projection * sin(angle*pi / 180)) %>% 
    summarize(avg_error = mean(sqrt((Petal.Length - proj_x)^2 + (Petal.Width - proj_y)^2)))
  
  return(iris_projection[1,1])
}

find_variance <- function(angle){
  iris_projection <- iris %>%
    mutate(projection = (Petal.Length - mean(iris$Petal.Length)) * cos(angle * pi / 180)  + (Petal.Width - mean(iris$Petal.Width)) * sin(angle * pi / 180),
           proj_x = mean(iris$Petal.Length) + projection * cos(angle * pi / 180), proj_y = mean(iris$Petal.Width) + projection * sin(angle*pi / 180))
  
  return(100 * var(iris_projection$projection) / (var(iris$Petal.Length) + var(iris$Petal.Width)))
}
```

```{r}
metrics <- tibble(angles = seq(from = 0, to = 45, by = 0.1)) %>% 
  add_column(error = sapply(angles, find_error)) %>% 
  add_column(variance = sapply(angles, find_variance))
```

```{r}
coeff = 1 / 100

metrics %>% 
  ggplot(aes(x = angles)) +
  geom_line(aes(y = error / coeff), color = 'red') + 
  geom_line(aes(y = variance), color = 'blue') +
  scale_y_continuous(name = "Variance",
    sec.axis = sec_axis(~.*coeff, name = "Reconstruction Error")
  ) +
  labs(x = "Angle") + 
  geom_vline(xintercept = angle, linetype = 'dashed') + 
  theme(
    axis.title.y = element_text(color = "blue"), 
    axis.title.y.right = element_text(color = "red")
  )
```