---
title: "Regression in R"
output: html_notebook
---

In this notebooks, we'll see how to build regression models in R. 

```{r}
library(tidyverse)
```


We'll start with the penguins dataset.
```{r}
penguins <- read_csv("../data/penguins.csv")
```

To build a linear regression model, you can use the lm function. We need to give a formula in the form of target ~ predictors and specify the dataset. Calling this function both creates and fits the model.

```{r}
penguins_lm <- lm("body_mass_g ~ flipper_length_mm", data = penguins)
```

Once fit, we can inspect the results using the summary function. 
```{r}
summary(penguins_lm)
```

If we need, we can extract the coefficients (as a named vector) using the coef function.
```{r}
coef(penguins_lm)
```

If we want to get estimates out of our model, we can use the predict function and specify the new data.
```{r}
predict(
  penguins_lm, 
  newdata = head(penguins)
  )
```

If we want a confidence or prediction interval around the estimate, we can specify that.
```{r}
predict(
  penguins_lm, 
  newdata = head(penguins),
  interval = 'confidence'  # can also do prediction interval
  )
```

```{r}
est_df <- tibble(
  flipper_length_mm = seq(from = min(penguins$flipper_length_mm, na.rm = TRUE), 
                          to = max(penguins$flipper_length_mm, na.rm = TRUE), 
                          length.out = 100)
    )

est_df <- est_df |> 
  bind_cols(predict(penguins_lm, newdata = est_df, interval = "confidence")) |> 
  rename(lwr_ci = lwr, upr_ci = upr) |> 
  bind_cols(predict(penguins_lm, newdata = est_df, interval = "prediction") |> 
              as_tibble() |> 
              select(lwr_pi = lwr, upr_pi = upr)
            )

penguins |> 
  drop_na(flipper_length_mm, body_mass_g) |> 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  geom_line(data = est_df, aes(x = flipper_length_mm, y = fit)) +
  geom_ribbon(data = est_df, aes(x = flipper_length_mm,
                                 y = fit,
                                 ymin = lwr_ci,
                                 ymax = upr_ci), alpha = 0.5) +
  geom_line(data = est_df, aes(x = flipper_length_mm, y = lwr_pi), linetype = 'dashed') +
  geom_line(data = est_df, aes(x = flipper_length_mm, y = upr_pi), linetype = 'dashed')
```


If we want to check for statistical significance, we can use the anova function. 
```{r}
reduced_model <- lm("body_mass_g ~ flipper_length_mm", data = penguins)
full_model <- lm("body_mass_g ~ flipper_length_mm + species", data = penguins)

anova(reduced_model, full_model)
```

In R, we can specify interaction terms the same as we did in Python.

```{r}
penguins_lm_interactions <- lm("body_mass_g ~ flipper_length_mm*species", 
                               data = penguins)

summary(penguins_lm_interactions)
```

## Logistic Regression

```{r}
field_goals <- read_csv("../data/fg.csv")
```

To fit a logistic regression model, we need to use the glm function.
```{r}
fg_lm <- glm("target ~ distance", 
             data = field_goals, 
             family = binomial)
```

Similar to the linear regression model above, we can look at the summary. 
```{r}
summary(fg_lm)
```

Note that the predict function will return the log-odds by default.
```{r}
predict(fg_lm, newdata = head(field_goals))
```

This can be changed by specifying the prediction type.
```{r}
predict(fg_lm, newdata = head(field_goals), type = "response")
```

```{r}
tibble(
  distance = seq(from = min(field_goals$distance),
                 to = max(field_goals$distance),
                 length.out = 100
                 ),
  estimate = predict(fg_lm,
                     newdata = tibble(distance = seq(from = min(field_goals$distance),
                 to = max(field_goals$distance),
                 length.out = 100
                 )),
                 type = "response")
) |> 
  ggplot(aes(x = distance, y = estimate)) +
  geom_line()
```

## Poisson Regression
```{r}
doctor_visits <- read_csv("../data/doctor_visits.csv")
```

Let's first look at the visits column, the column of interest.
We can see that overall, it is right-skewed with a large number of zeros.
```{r}
doctor_visits |> 
  ggplot(aes(x = visits)) + 
  geom_bar()
```

We could start with a constant-only model.
```{r}
pr_hosp_const <- glm('visits ~ 1',
               data = doctor_visits,
               family = poisson)
summary(pr_hosp_const)
```

Let's see what it estimates for the mean number of the distribution:
```{r}
mu = exp(coef(pr_hosp_const))
```

And plot the result.
```{r}
x <- 0:5
y <- dpois(x, lambda = mu)

tibble(x = x, y = y) |> 
  ggplot(aes(x = x, y = y)) + 
  geom_col()
```

It looks like it does okay, but could we do better by including another covariate? We'll use age.
```{r}
doctor_visits <- doctor_visits |> 
  mutate(age = 100*age)
```

If we want to see how the average number of visits per age, we could use the ntile function, which will divide our data into equal-sized groups.
```{r}
doctor_visits |> 
  mutate(group = ntile(age, n = 5)) |> 
  group_by(group) |> 
  summarize(
    mean_age = mean(age),
    mean_visits = mean(visits)
    )
```

```{r}
doctor_visits |> 
  mutate(group = ntile(age, n = 5)) |> 
  group_by(group) |> 
  ggplot(aes(x = visits)) +
  geom_bar() +
  facet_wrap(~group)
```

```{r}
pr_hosp <- glm('visits ~ age',
               data = doctor_visits,
               family = poisson)
```

```{r}
summary(pr_hosp)
```

```{r}
est_df <- tibble(
  age = seq(from = min(doctor_visits$age, na.rm = TRUE), 
                          to = max(doctor_visits$age, na.rm = TRUE), 
                          length.out = 100)
    )

est_df <- est_df |> 
  bind_cols(predict(pr_hosp, newdata = est_df, type = "response")) |> 
  rename("Estimated Mean Number of Visits" = "...2")
            
est_df |> 
  ggplot(aes(x = age, y = `Estimated Mean Number of Visits`)) +
  geom_line()
           
```

You can adjust the value of age to see how the estimated distribution changes
```{r}
age <- seq(from = 20, to = 70, by = 10)
mu <- predict(pr_hosp, newdata = tibble(age = age), type = "response")

map(age, 
    \(x) tibble(age = x, 
                visits = 0:5, 
                probability = dpois(0:5, 
                                    lambda = predict(pr_hosp, 
                                                     newdata = tibble(age = x), 
                                                     type = "response")
                                    )
                )
    ) |> 
  bind_rows() |> 
  ggplot(aes(x = visits, y = probability)) +
  geom_col() +
  facet_wrap(~age)
```

Example 2: Including an offset term.
```{r}
mlb <- read_csv("../data/mlb_2021.csv") |> 
  rename(runs = R, games = G)
```

```{r}
pr_mlb <- glm("runs ~ OBP",
              data = mlb,
              family = poisson,
              offset = log(games))
```

```{r}
summary(pr_mlb)
```

```{r}
est_df <- tibble(
  OBP = seq(from = min(mlb$OBP, na.rm = TRUE), 
                          to = max(mlb$OBP, na.rm = TRUE), 
                          length.out = 100),
  games = 1
    )

est_df <- est_df |> 
  bind_cols(predict(pr_mlb, newdata = est_df, type = "response")) |> 
  rename("Estimated Mean Runs Per Game" = "...3")

mlb |> 
  ggplot(aes(x = OBP, y = `R/G`)) +
  geom_point() +
  geom_line(data = est_df, aes(x = OBP, y = `Estimated Mean Runs Per Game`))
```
